\Large Inferential Analysis of the Supreme Court Citation Network

### 2018-03-26 Tex/paper.tex

Let $C_t$ be the case-to-case adjacency matrix at time $t$ such that $C_t^{ij} \in \{0,1\}$ is a binary indicator of whether case $i$ cites $j$. Furthermore, let
$$\mathcal{C}_t =\{C_{\leq t} \in \{0,1\}^{(N_t \cdot (N_t-1))/2}: C_{\leq t}^{ij} \in \{0,1\} \}$$ 
be the set of all possible adjacency matrices among $N_t$ cases. $N_t$ denotes the the number of cases in the network at time $t$. Note that the cardinality of $\mathcal{C}_t$ increases exponentially for every newly added case. The probability of observing $C_t$ given past citations $C_{<t}$, where $C_{\leq t}$ is the network up to time $t$ is defined as
\begin{equation}
P(C_t | C_{<t}, \theta)=\cfrac{\exp\big(\theta^T\cdot h\big(C_{\leq t})\big)\big)}{\sum_{Z_t^*\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)\big)}
\label{ercm}
\end{equation}
where $\theta \in \mathbb{R}^q$ is a $q$-dimensional vector of parameters,  $h: \mathcal{C}_t \to \mathbb{R}^q$ is a q-dimensional vector of different statistics and $\kappa(\theta) := \sum_{Z^*_t\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)$ is a normalization constant that ensures that Equation (\ref{ercm}) defines a probability function on $\mathcal{C}_t$.

% Network Statistics
The c-TERGM is specified through 
the decision regarding which network statistics $h(\cdot)$ to incorporate. We include the following statistics for the Supreme Court citation network, which are derived from the interdependence hypotheses described in the previous section:
$$h_{edges}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{ij} C_{\leq t}^{ij}, $$
the number of citations.  $h_{edges}$ performs the function of an intercept, and models the expected value of any given edge.
$$h_{outstar}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{i=1} {\sum_{j \neq i} C_{\leq t}^{ij} \choose 2} $$
the number of out-two-stars. An out-two-star is a configuration in which one node sends to two other nodes. The number of these configurations grows quadratically as the origins of ties concentrate on a few highly active/social senders. The out-two-stars configuration is commonly used to model activity (sender activation) in ERGMs.
$$h_{instar}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{j=1} {\sum_{i \neq j} C_{\leq t}^{ij} \choose 2} $$
the number of in-two-stars. An in-two-star is a configuration in which one node receives ties from two other nodes. The number of these configurations grows quadratically as the destinations of ties concentrate on a few highly popular recipients. The out-two-stars configuration is commonly used to model popularity in ERGMs.
\begin{align*}
h_{triangle}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to & \sum_{j<i<k} C^{ij}\cdot C^{jk} \cdot C^{ki} + C^{ji}\cdot C^{jk} \cdot C^{ki} + C^{ij}\cdot C^{kj} \cdot C^{ki} \\ & + C^{ji}\cdot C^{kj} \cdot C^{ki} + C^{ij}\cdot C^{jk} \cdot C^{ik} + C^{ji}\cdot C^{jk} \cdot C^{ik}  \\ & + C^{ij}\cdot C^{kj} \cdot C^{ik} + C^{ji}\cdot C^{kj} \cdot C^{ik},  
\end{align*}
The number of triangles in the network.  Triangles measure triad closure, as discussed above. $$h_{covariate}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{ij} C_{\leq t}^{ij}X_{ij}, $$ the effect of an exogenous covariate ($X$). We include several exogenous covariates based on this standard statistic formulation. These covariates are discussed below.

The structure of this model is very similar to a conventional (T)ERGM setup. The main difference is that  $\mathcal{C}_t$ excludes adjacency matrices in which there are loops that include edges sent at different time points. Loops can exist in the Supreme Court citation network, but only among cases decided in the same term. The normalizing constant in the denominator of the probability function (i.e., Equation \ref{ercm}) is computationally intractable. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with ERGM family models. Methods of Monte Carlo approximation are well established \citep{hunter2006inference,van2009framework,hummel2012improving}, but with almost $10,000$ vertices, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. To estimate the parameters of the c-TERGM, we rely on two features of the model and data that facilitate the application of an approximate inference method that is faster than Monte Carlo approximation---maximum pseudolikelihood. First, our sample size is very large. Second, under the c-TERGM we assume that citations are formed in each time period conditional upon the citations that exist in the past.  We estimate the c-TERGM via bootstrapped pseudolikelihood \citep{desmarais2012statistical,desmarais2010consistent}.

@@

\begin{table}[]
\centering
\begin{tabular}{llll}
1 & Criminal Procedure & 8  & Economic Activity    \\
2 & Civil Rights       & 9  & Judicial Power       \\
3 & First Amendment    & 10 & Federalism           \\
4 & Due Process        & 11 & Interstate Relations \\
5 & Privacy            & 12 & Federal Taxation     \\
6 & Attorneys          & 13 & Miscellaneous        \\
7 & Unions             & 14 & Private Action      
\end{tabular}
\caption{Code Book for the variable \textit{Issue Area}. This information originates from the Supreme Court Database code book.}
\label{issue_area_coding}
\end{table}

@@

\begin{table}[htp]
\centering
\begin{tabular}{|
>{\columncolor[HTML]{C0C0C0}}l |l|l|l|}
\hline
{\color[HTML]{333333} } & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Terms} & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Total Number Cases} & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Cases/Term} \\ \hline
CE Hughes*              & 1937 - 1941                                          & 628                                                               & 125.6                                                     \\ \hline
HF Stone                & 1942 - 1946                                          & 756                                                               & 151.2                                                     \\ \hline
FM Vinson               & 1946 - 1953                                          & 789                                                               & 98.63                                                      \\ \hline
E Warren                & 1954 - 1969                                          & 2149                                                              & 126.41                                                     \\ \hline
WE Burger               & 1970 - 1986                                          & 2805                                                             & 155.83                                                     \\ \hline
W Rehnquist            & 1987 - 2005                                          & 2022                                                              & 106.42                                                      \\ \hline
J Roberts **           & 2006 - 2015                                          & 796                                                              & 79.6                                           \\ \hline
\end{tabular}
\caption[caption]{For the time range of interest (1937 - 2015) this table displays the chief justices, the time range they served as chief justice, the number of cases in their time range as well as the average number of cases per year.\\\hspace{\textwidth} * CE Hughes served as chief justice from 1930 - 1941. \\\hspace{\textwidth} ** J Roberts still serves as chief justice (3/2018).}
\label{tab:chiefs}
\end{table}

@@

\begin{figure}[htp]
\includegraphics[scale=0.35,clip=true,trim=.5cm 0cm 0cm 2cm]{images/citations1}
\caption{ Supreme Court Citation Network, 1937?2001. Nodes are Supreme Court cases, with size based on incoming citations. Salient cases ({\em Oxford Guide}) are in yellow. Each network layer contains cases decided under a different chief justice. Because our data is temporally bound, cases for the Hughes and Rehnquist courts are not complete. }
 \label{fig:networkviz}
\vspace{-.25cm}
\end{figure}

@@

 (e) Overruled Cases & (f) Difference in Years \\
\includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/overruled_coef_trend_linear.pdf} & \includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/yeardiff_coef_trend_linear.pdf} \\

   (g) Reciprocity & (h) Transitivity \\
\includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/mutual_coef_trend_linear.pdf} & \includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/triangle_coef_trend_linear.pdf} \\

    (i) Activity & (j) Popularity \\
\includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/o2star_coef_trend_linear.pdf} & \includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/i2star_coef_trend_linear.pdf} \\

\end{tabular}
\caption{Trending effects from the c-TERGM. The solid lines plot the point estimates, and the shaded areas span 95\% confidence intervals. }
 \label{fig:coeftrends}
\vspace{-.25cm}
\end{figure}

>>>>

Let $C_t$ be the case-to-case adjacency matrix at time $t$ such that $C_t^{ij} \in \{0,1\}$ is a binary indicator of whether case $i$ cites $j$. Furthermore, let
$$\mathcal{C}_t =\{C_{\leq t} \in \{0,1\}^{(N_t \cdot (N_t-1))/2}: C_{\leq t}^{ij} \in \{0,1\} \}$$ 
be the set of all possible adjacency matrices among $N_t$ cases at time $t$. $N_t$ denotes the the number of cases in the network at time $t$. Note that the cardinality of $\mathcal{C}_t$ increases exponentially for every newly added case. The probability of observing $C_t$ given past citations $C_{<t}$, where $C_{\leq t}$ is the network up to time $t$ is defined as
\begin{equation}
P(C_t | C_{<t}, \theta)=\cfrac{\exp\big(\theta^T\cdot h\big(C_{\leq t})\big)\big)}{\sum_{Z_t^*\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)\big)}
\label{ercm}
\end{equation}
where $\theta \in \mathbb{R}^q$ is a $q$-dimensional vector of parameters,  $h: \mathcal{C}_t \to \mathbb{R}^q$ is a q-dimensional vector of different statistics and $\kappa(\theta) := \sum_{Z^*_t\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)$ is a normalization constant that ensures that Equation (\ref{ercm}) defines a probability function on $\mathcal{C}_t$.

% Network Statistics
The c-TERGM is specified through 
the decision regarding which network statistics $h(\cdot)$ to incorporate. We include the following network statistics for the Supreme Court citation network, which are derived from the interdependence hypotheses described in the previous section:
$$h_{edges}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{ij} C_{\leq t}^{ij}, $$
the number of citations.  $h_{edges}$ performs the function of an intercept, and models the expected value of any given edge.
$$h_{outstar}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{i=1} {\sum_{j \neq i} C_{\leq t}^{ij} \choose 2} $$
the number of out-two-stars. An out-two-star is a configuration in which one node sends to two other nodes. The number of these configurations grows quadratically as the origins of ties concentrate on a few highly active/social senders. The out-two-stars configuration is commonly used to model activity (sender activation) in ERGMs.
$$h_{instar}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{j=1} {\sum_{i \neq j} C_{\leq t}^{ij} \choose 2} $$
the number of in-two-stars. An in-two-star is a configuration in which one node receives ties from two other nodes. The number of these configurations grows quadratically as the destinations of ties concentrate on a few highly popular recipients. The out-two-stars configuration is commonly used to model popularity in ERGMs. 
\begin{align*}
h_{triangle}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to & \sum_{j<i<k} C^{ij}\cdot C^{jk} \cdot C^{ki} + C^{ji}\cdot C^{jk} \cdot C^{ki} + C^{ij}\cdot C^{kj} \cdot C^{ki} \\ & + C^{ji}\cdot C^{kj} \cdot C^{ki} + C^{ij}\cdot C^{jk} \cdot C^{ik} + C^{ji}\cdot C^{jk} \cdot C^{ik}  \\ & + C^{ij}\cdot C^{kj} \cdot C^{ik} + C^{ji}\cdot C^{kj} \cdot C^{ik},  
\end{align*}
The number of triangles in the network. Triangles measure triad closure, as discussed above.      
$$h_{reciprocity}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{ij} C_{\leq t}^{ij}C_{\leq t}^{ji}, $$
The number of reciprocal ties in the network. Even though the reciprocity of ties is not a meaningful network statistic for most citation networks, it may still play a role in the Supreme Court citation network. This statistic comes with the restriction that it only can appear among cases that entered the network at the same time.

$$h_{covariate}:\mathcal{C}_t \to \mathbb{R}~~~, ~~~C_{\leq t} \to \sum_{ij} C_{\leq t}^{ij}X_{ij}, $$ the effect of an exogenous covariate ($X$). We include several exogenous covariates based on this standard statistic formulation. These covariates are discussed below.

The structure of this model is very similar to a conventional (T)ERGM setup. The main difference is that  $\mathcal{C}_t$ excludes adjacency matrices in which there are loops that include edges sent at different time points. Loops can exist in the Supreme Court citation network, but only among cases decided in the same term. Furthermore, a case $i$ can only cite a case $j$ if $j$ entered the network before or at the same time as $i$ did. The normalizing constant in the denominator of the probability function (i.e., Equation \ref{ercm}) is computationally intractable. This means that straightforward methods of maximum likelihood estimation (MLE) are infeasible with ERGM family models. Methods of Monte Carlo approximation are well established \citep{hunter2006inference,van2009framework,hummel2012improving}, but with almost $10,000$ vertices, even the methods of Monte Carlo MLE are computationally expensive \citep{schmid2017exponential}. To estimate the parameters of the c-TERGM, we rely on two features of the model and data that facilitate the application of an approximate inference method that is faster than Monte Carlo approximation---maximum pseudolikelihood. First, our sample size is very large. Second, under the c-TERGM we assume that citations are formed in each time period conditional upon the citations that exist in the past.  We estimate the c-TERGM via bootstrapped pseudolikelihood \citep{desmarais2012statistical,desmarais2010consistent}.

@@

\begin{table}[]
\centering
\begin{tabular}{llll}
1 & Criminal Procedure & 8  & Economic Activity    \\
2 & Civil Rights       & 9  & Judicial Power       \\
3 & First Amendment    & 10 & Federalism           \\
4 & Due Process        & 11 & Interstate Relations \\
5 & Privacy            & 12 & Federal Taxation     \\
6 & Attorneys          & 13 & Miscellaneous        \\
7 & Unions             & 14 & Private Action      
\end{tabular}
\caption{Assigned numbers for the variable \textit{Issue Area}. This information originates from the Supreme Court Database code book.}
\label{issue_area_coding}
\end{table}

@@

\begin{table}[htp]
\centering
\begin{tabular}{|
>{\columncolor[HTML]{C0C0C0}}l |l|l|l|}
\hline
{\color[HTML]{333333} } & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Terms} & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Total Number Cases} & \cellcolor[HTML]{C0C0C0}{\color[HTML]{333333} Cases/Term} \\ \hline
CE Hughes*              & 1937 - 1941                                          & 628                                                               & 125.6                                                     \\ \hline
HF Stone                & 1942 - 1946                                          & 756                                                               & 151.2                                                     \\ \hline
FM Vinson               & 1946 - 1953                                          & 789                                                               & 98.63                                                      \\ \hline
E Warren                & 1954 - 1969                                          & 2149                                                              & 126.41                                                     \\ \hline
WE Burger               & 1970 - 1986                                          & 2805                                                             & 155.83                                                     \\ \hline
W Rehnquist            & 1987 - 2005                                          & 2022                                                              & 106.42                                                      \\ \hline
J Roberts **           & 2006 - 2015                                          & 796                                                              & 79.6                                           \\ \hline
\end{tabular}
\caption[caption]{For the time range of interest (1937 - 2015) this table displays the chief justices, the time range they served as chief justice, the number of cases in their time range as well as the average number of cases per year.\\\hspace{\textwidth} * CE Hughes served as chief justice from 1930 - 1941. \\\hspace{\textwidth} ** J Roberts still serves as chief justice (retrieved 3/2018).}
\label{tab:chiefs}
\end{table}

@@

\begin{figure}[htp]
\includegraphics[scale=0.35,clip=true,trim=.5cm 0cm 0cm 2cm]{images/citations1}
\caption{ Supreme Court Citation Network, 1937-2001. Nodes are Supreme Court cases, with size based on incoming citations. Salient cases ({\em Oxford Guide}) are in yellow. Each network layer contains cases decided under a different chief justice. Because our data is temporally bound, cases for the Hughes and Rehnquist courts are not complete. }
 \label{fig:networkviz}
\vspace{-.25cm}
\end{figure}

\begin{itemize}
\item update Figure 2 to time period 1937 - 2015
\end{itemize}

@@

 (e) Difference in Years & (f) Difference in Years$^2$ \\
\includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/yeardiff_coef_trend_linear.pdf} & \includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/yeardiffsquare_coef_trend_linear.pdf} \\

 (g) Overruled Cases &  \\
\includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/overruled_coef_trend_linear.pdf} &  \\

 \end{tabular}
\caption{Trending effects for exogenous variables from the c-TERGM. The solid lines plot the point estimates, and the shaded areas span 95\% confidence intervals. }
 \label{fig:coeftrends_exo}
\vspace{-.25cm}
\end{figure}

\begin{figure}[H]
 \begin{tabular}{cc}

   (h) Reciprocity & (i) Transitivity \\
\includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/mutual_coef_trend_linear.pdf} & \includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/triangle_coef_trend_linear.pdf} \\

    (j) Activity & (k) Popularity \\
\includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/o2star_coef_trend_linear.pdf} & \includegraphics[width = 0.475\textwidth, trim= 0.1cm 1cm 0.5cm .45cm,clip=true]{images/i2star_coef_trend_linear.pdf} \\

\end{tabular}
\caption{Trending effects for endogenous statistics from the c-TERGM. The solid lines plot the point estimates, and the shaded areas span 95\% confidence intervals. }
 \label{fig:coeftrends_endo}
\vspace{-.25cm}
\end{figure}
